On change of priors in one-vs-one classification methods

### 2025-05-07 paper.tex
Trying to make generated graphs
Fix build
Starting rewrite of proofs
Fix names of theory sections

\title{On probabilistic one-vs-one classification}

@@

\subsection{Bayes classifier}

An exact solution to this problem is provided by the Bayes classifier. To simplify the discussion we will make the following assumption.

@@

\begin{prop} \label{prop:binary2multi}
The system of  equations \eqref{eq:bt1} in unknowns $p_i^\Pi(x)$ has a unique solution which is the multi-class Bayes classifier.
\end{prop}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

\subsection{One-vs-one classification }

In practice one does not have know true probability distributions and hence cannot construct Bayes classifier. However, given a sample from the distribution, we can train binary classifiers to provide approximations $r_{ij}(x)$ to pairwise Bayes classifiers $p_{ij}^\Pi(x)$. The goal of probabilistic one-vs-one classification is to deduce for any $x$ in the feature space $X$ an estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\Pi(x)$. One does it by finding an approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

@@

In order to obtain a regular system of equations for  larger values of $K$ , the resulting classifiers must represent a connected tree within the complete graph on all classes. Any arboreal method is non-canonical, because there is no natural way to choose a tree in a complete graph.

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\boldsymbol{s}_i$ the prediction of the arboreal method, where the underlying tree is a star centered at class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\boldsymbol{s}_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 \boldsymbol{s}_1(x) + \ldots + \hat p_K \boldsymbol{s}_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${{T}}_i$.

\subsection{Bayes covariant methods}

There are many Bayes covariant methods. First examples are provided by arboreal coupling methods.

\begin{prop}
Any arboreal method $\boldsymbol{t}$ is Bayes covariant.
\end{prop}
\begin{proof}
We start by defining property $BC_{ij}$ (short for Bayes covariant when restricted to classes $i$ and $j$) of a coupling method. A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal.

@@

@@

Let $\pi$ be a change of priors from $(P_1, \ldots, P_K)$  to $(P'_1, \ldots, P'_K)$. Let $o_m$ denotes the odds of classes $C_i$ and $C_j$ predicted by coupling method $\boldsymbol{v}_m$. Then the odds of $C_i$ and $C_j$ predicted by $(\bigoplus_i a_i \boldsymbol{v}_i)^\pi$ are 
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)^\pi(\boldsymbol{R}) & = \frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R}) \\
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & =  \biggl(\bigoplus_m a_m \boldsymbol{v}_m^\pi\biggr)(\boldsymbol{R}) \\
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

@@

Let us point out one consequence of the theorem. We could define another ensemble $\boldsymbol{E}$ by averaging the predictions of $\boldsymbol{s}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$:
\[
\boldsymbol{E} = \frac13 \boldsymbol{s}_1(x) + \frac13 \boldsymbol{s}_2(x) + \frac 13 \boldsymbol{s}_3(x).
\]
Note that here we use the plus symbol to denote the vector addition of vectors representing $K$-class probability distributions. This is a different classifier from the one given by  \eqref{eq:bc1} (see Appendix \ref{app:explicit}), and since it is symmetric, it cannot be Bayes covariant.

@@

\begin{figure}[!ht]
\includegraphics{graph/exp2-summary.pdf}
\caption{Results of grid search for parametric Bayes covariant methods matching performance of Wu-Lin-Weng's method}
\label{fig:par-bc}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics{graph/exp2-detail.pdf}
	\caption{Color map indicates the value of Brier score computed for an ensemble of coupling methods on the testing dataset. The vertices triangle indicate the arboreal coupling methods, and the black circle the location of the optimum.}
	\label{fig:score}
\end{figure}

>>>>

\title{On change of priors in one-vs-one classification methods}

@@

\subsection{One-vs-on classification}

An exact solution to a classification problem is provided by the Bayes classifier. To simplify the discussion we will make the following assumption.

@@

\begin{prop}[Fundamental theorem of one-vs-one classification]
	 \label{prop:binary2multi}
	 Knowledge of binary Bayes classifiers $p_{ij}^\Pi$ determines the multiclass Bayes classifier $\boldsymbol{p}^\Pi$.
\end{prop}
\begin{proof}
	Indeed, from \eqref{eq:bt1} we deduce the odds of classes $C_j$ to $C_i$
	\begin{align}
	\frac{p_j^\Pi(x)}{p_j^\Pi(x)} = \frac{1}{p^\Pi_{ij}(x)}- 1.
	\end{align}
	Knowing the odds of $C_2$ to $C_1$, $C_3$ to $C_2$ up to $C_{K}$ to $C_{K-1}$ together with the requirement that probabilities sum to one uniquely determines the whole vector $\boldsymbol{p}^\Pi(x)$.
\end{proof}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

\subsection{Coupling methods}

In practice one does not have know true probability distributions and hence cannot construct Bayes classifies. However, given a sample from the distribution, we can train binary classifiers to provide approximations $r_{ij}(x)$ to pairwise Bayes classifiers $p_{ij}^\Pi(x)$. The goal of probabilistic one-vs-one classification is to deduce for any $x$ in the feature space $X$ an estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\Pi(x)$. One does it by finding an approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed to construct an estimator. A method to solve Bradley-Terry equations is called a \emph{coupling method}. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

@@

For $K>3$ the three methods can be generalized to corresponds to the trees having the graph structure  of a star centered at class $C_i$ for $i=1, \ldots, K$.

More generally, to obtain a regular system of equations for  any of $K$ , it is necessary that the tree corresponding to an arboreal method forms a spanning tree. Note that any arboreal method is non-canonical, because there is no natural way to choose a tree in a complete graph. In fact, a stronger statement is true -- no arboreal method is symmetric.
%However, we will see in Section \ref{sec:bc2} that by ``averaging'' several stars we can obtain a canonical

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\boldsymbol{s}_i$ the prediction of the arboreal method, where the underlying tree is a star centred at the class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\boldsymbol{s}_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 \boldsymbol{s}_1(x) + \ldots + \hat p_K \boldsymbol{s}_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${\boldsymbol{s}}_i$.

\subsection{Bayes covariant methods} \label{sec:bc2}

There are many Bayes covariant methods. In this section we will show that 
\begin{itemize} 
	\item arboreal coupling methods are Bayes covariant,
	\item Bayes covariant methods can be ensembled,
	\item describe all Bayes covariant methods for $K=3$.
\end{itemize}

We start by defining the property $BC_{ij}$ (the notation stands for \emph{Bayes covariant when restricted to classes} $i$ and $j$) of a coupling method. A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal.

@@

Now we are ready to prove the our first result.

\begin{prop}
Any arboreal method is Bayes covariant.
\end{prop}
\begin{proof}

@@

Let $\pi$ be a change of priors from $(P_1, \ldots, P_K)$  to $(P'_1, \ldots, P'_K)$. Let $o_m$ denotes the odds of classes $C_i$ and $C_j$ predicted by coupling method $\boldsymbol{v}_m$. Then the odds of $C_i$ and $C_j$ predicted by $(\bigoplus_i a_i \boldsymbol{v}_i)^\pi$ are 
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)^\pi(\boldsymbol{R}) & = \frac{P'_i}{P_i} \frac{P_j}{P'_j} \biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R}) \\
\end{align}
so the odds are

\begin{align}
	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

Let us now use Bayes covariance of $\boldsymbol{v}_i$. 
We have
\begin{align}
	\biggl(\bigoplus_m a_m \boldsymbol{v}_m\biggr)(\boldsymbol{R^\pi}) & =  \biggl(\bigoplus_m a_m \boldsymbol{v}_m^\pi\biggr)(\boldsymbol{R}) \\
%	&= \frac{P'_i}{P_i} \frac{P_j}{P'_j} \prod_m o_m^{a_m}
\end{align}

@@

Let us point out one consequence of the theorem. We could define another ensemble $\boldsymbol{E}$ by averaging the predictions of $\boldsymbol{s}_1, \boldsymbol{s}_2, \boldsymbol{s}_3$:
\[
\boldsymbol{E} = \frac13 \boldsymbol{s}_1(x) + \frac13 \boldsymbol{s}_2(x) + \frac 13 \boldsymbol{s}_3(x).
\]
Note that here we use the plus symbol to denote the vector addition of vectors representing $K$-class probability distributions. This is a different classifier from the one given by  \eqref{eq:bc1} (see Appendix \ref{app:explicit}), and since it is symmetric, it cannot be Bayes covariant.

@@

\begin{figure}[!ht]
\includegraphics{graphs/exp2-summary.pdf}
\caption{Results of grid search for parametric Bayes covariant methods matching performance of Wu-Lin-Weng's method}
\label{fig:par-bc}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics{graphs/exp2-detail.pdf}
	\caption{Color map indicates the value of Brier score computed for an ensemble of coupling methods on the testing dataset. The vertices triangle indicate the arboreal coupling methods, and the black circle the location of the optimum.}
	\label{fig:score}
\end{figure}
